Distributional Trees and Forests for Circular Data

### 2019-09-19 pkg/circtree/inst/wind_paper/draft/draft.Rnw
small revisions on methodology section
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2764 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: f2129efc5b46c474e76a5dc39bf2784b5d30a4f9

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle applying the modulo operation on the 
interval length $2\pi$ \citep{Mardia+Jupp:2009}.
In that way, for example the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in [0, 2\pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
One way how to deal with these problems is to apply a tree-structured model.

@@

In particular, in each node the crucial decision on how and where to split the data is based on model scores which are obtained by evaluating the score function $s$ at the individual observations and 
parameter estimates. For the von Mises distribution and a data set of $n$ observations
this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
a given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

Once a circular distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node depending on the values of the covariates and in that way to the corresponding estimated parameter pair $(\hat{\mu}, \hat{\kappa})$ of this node. Therefore, the same parameter pair is estimated for all observations
ending up in the same terminal node. This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
from the learning sample and the fitted model. This can also be expressed by employing weights

\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

>>>>

\section{Probabilistic Circular Modeling}
\label{sec:prob_circ}
Probabilistic modeling of circular data requires the selection of a proper circular
probability distribution, i.e., a probability distribution which accounts for the periodicity
of circular data.
Generally, this feature can be obtained by ``wrapping'' the probability density function
of any continuous distribtuion around the unit circle \citep{Mardia+Jupp:2009}, for example by
applying the modulo operation on the interval length $2\pi$.
In that way, the wrapped Cauchy distribution or the wrapped normal distribution can be employed to model symmetric unimodal circular data.
A close approximation to the wrapped normal distribution that is mathematically simpler and hence easier 
to use \citep{Fisher:1993} is provided by the von Mises distribution,
a purely circular distribution which is also known as ``the circular normal distribution''
and commonly applied for probabilistic modeling of circular data.
Based on a location parameter $\mu \in [0, 2\pi]$ and a concentration parameter $\kappa > 0$ 
the density of the von Mises distribution for an observation $y \in [0, 2\pi]$ is given by:
\begin{equation}
  f_\mathrm{vM}(y; \mu, \kappa) = \frac{1}{2 \pi I_0(\kappa)}~e^{ \kappa \cos(y - \mu)}\label{schlosser:equ_vm}
\end{equation}
where $I_0(\kappa)$ is the modified Bessel function of the first kind and order $0$
\citep[see, e.g.,][for a more detailed overview]{Jammalamadaka+Sengupta:2001}.

@@

\subsection{Motivation}
A commonly used approach to including covariates as regressors in distributional models is 
provided by generalized additive models for location, scale, and shape 
\citep[GAMLSS, ][]{Rigby+Stasinopoulos:2005}.
There, each distribution parameter is linked to a linear predictor by an additive function.
While this strategy works well for smooth effects difficulties can arise in case of 
non-additive changes. Moreover, specifying a model properly can be very challenging,
particularly for a high number of covariates and no information about possible interactions.
An alternative approach that can deal with these problems is to apply a tree-structured model.

@@

In particular, in each node the crucial decision on how and where to split the data is based on model scores which are obtained by evaluating the score function $s$ at the individual observations and 
parameter estimates. For the von Mises distribution and a data set of $n$ observations
this yields an $n \times 2$ matrix that can be employed as a kind of residual, capturing how well
each given observation conforms with the estimated location $\hat{\mu}$ and precision $\hat{\kappa}$, 
respectively.
To capture dependence on covariates, the association between the model's scores and each available
covariate is assessed using either a parameter instability test (MOB) or a permutation test (CTree).
In each partitioning step, the covariate with the highest significant association (i.e., lowest
significant $p$-value, if any) is selected for splitting the data. The corresponding split point
is chosen either by optimizing the log-likelihood (MOB) or a two-sample test statistic (CTree)
over all possible partitions.
%and the covariate corresponding to the strongest change is selected as split varialbe.
%Next a split point is chosen within the selected covariate by maximizing a partitioned likelihood.
This procedure is repeated recursively until there are no significant parameter instabilities or until another stopping criterion is met (e.g., subgroup size or tree depth).

@@

Once a circular distributional tree model is fit it can be employed to obtain probabilistic predictions
for a possibly new set of observations of the covariates $\bm{z} = (z_1, \ldots, z_m)$.
Starting at the root node the tree structure leads the observation to a terminal node, depending on the values of the covariates, where the parameter pair $(\hat{\mu}, \hat{\kappa})$ is estimated for
the corresponding subset of learning observations. 
This can also be expressed by employing weights which indicate whether the $i$-th learning observation and the observation $\bm{z}$ belong to the same terminal
node:
\begin{equation}
w^{\text{tree}}_i(\bm{z}) = \sum_{b=1}^B \mathbf{1}((\bm{z}_i \in \mathcal{B}_b) \land (\bm{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function and $\mathcal{B}_b$ is the $b$-th out of $B$ segments
partitioning the covariate space in disjoint subsets. 
In that way the estimated parameter pair $(\hat{\mu},\hat{\kappa})(\bm{z})$ specifying the predicted
von Mises distribution for a given $\bm{z}$ is obtained by
\begin{equation}
(\hat{\mu},\hat{\kappa})(\bm{z}) = \argmax_{\mu,\kappa} \sum_{i=1}^n w^{\text{tree}}_i(\bm{z}) \cdot \ell(\mu,\kappa; y_i).
\end{equation}

Therefore, the same parameter pair is estimated for all observations
belonging to the same terminal node. This facilitates the application as the parameter estimates do not need to be recalculated for each (new) observation via maximum likelihood but can be extracted directly
from the learning sample and the fitted model.
